Reciprocal Tolerance: Why States Allow Their Adversaries to Practice Espionage with Minimal Consequences

### 2020-11-08 article.tex
Remove "p" from page numbers

\maketitle
\newpage
\input{abstract.tex}
\newpage

In 2015, federal authorities discovered a massive data breach inside the United States Office of Personnel Management (OPM). Hackers had, over the course of months, downloaded reams of sensitive data, including a cache of applications for federal security clearance. These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment. \enquote{[I]f you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States. Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} In the case of the OPM hack, a vivid imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018. See \cite{sanger_trump_2018}}

@@

\subsection{Defining espionage norms}
There are few extant definitions of peacetime espionage, and no satisfying ones. The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost never defined by legal codes, or interpreted by case law. Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[p.~901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper. See \cite[p.~600]{radsan_unresolved_2007}, \cite[p.~2]{kapp_spying_2007})} Excluding these activities is important: assassinations and similar covert operations are not considered espionage, so they do not typically receive the protections of espionage norms.

@@

\subsection{Extant explanations for espionage norms}
Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage. In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments. See \cite{wright_essays_1962}} Espionage is universally illegal but every state acknowledges that allies and adversaries alike are guaranteed to engage in it.\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} Because of that tension, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact. Jared Beim concludes that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A. John Radsan employs a more philosophical approach. Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but the international system leaves this tension in a state of salutary neglect. Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.

Another branch of scholarship sidesteps the legal question in favor of a \enquote{functional theory} of IR. This approach is more informative because it attempts to explain why states tolerate espionage rather than creating \emph{post hoc} justifications for it. A functional approach, for instance, might highlight the role of espionage in treaty verification, the way Christopher Baker does when he argues \enquote{the advantages that espionage offers over legally-binding verification and assurance regimes tip the scales in favor of functional cooperation.}\footcite{baker_tolerance_2004} Baker correctly notes that the additional assurance provided by espionage enables states to enter into riskier agreements that they would otherwise have avoided. His analysis is limited, however, by its focus on enabling specific policy outcomes, like treaties. It doesn't resolve the gap between the absolute illegality of espionage in domestic law and its widespread practice internationally, nor does address why espionage would be tolerated in cases where no tangible outcome, like a treaty, is forthcoming. Responding to Baker, Radsan notes that one-off deals and treaties \enquote{are a very slow and indirect way to achieve international consensus on the legality of espionage. The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007}

@@

Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage. The results of intelligence operations can have devastating, even existential effects on national security. In recent years China has relentlessly targeted American military secrets. A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies], all with little to no adverse consequences to the thieves.}\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.

@@

\subsection{A model for maintaining espionage norms}
I propose an iterated decision-making model to describe how states cooperate to preserve the liminal legal status of espionage, based on the same tit-for-tat strategy associated with diplomatic expulsions. In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.\footcite[p.~8]{axelrod_effective_1980} By employing this strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same. Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind. Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal removal of diplomats to stop.

@@

To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure. Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond by not building up arms, and resolve the security dilemma for both of them. The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky. The state could be wrong that its adversary is a security seeker---what Glaser calls a \enquote{greedy state}---in which case its optimistic overtures of cooperation will have exposed it to greater risk of attack. When states are unwilling to make those signaling measures because they fear compromising their own security, then a security dilemma results.

Espionage creates the conditions that make solving the security dilemma possible. Consider a situation in which one state is a genuine security-seeker and so is its adversary. The state conducts espionage on its adversary, and the resulting intelligence increases its internal estimate of how likely the adversary is to be a security-seeker. As a result, the state decides that some form of arms control is no longer too risky to attempt in order to signal its benign intentions. Because the state receiving the signal is also a security-seeker, this is a welcome development; it could be empowered to signal its benign intentions as well, based on its own intelligence. For both states, this is the optimal outcome, contributing to both information variables that Glaser identifies as key to the security dilemma: \enquote{the state’s estimate of the adversary’s motives, and the state’s estimate of the adversary’s estimate of the state’s motives.}\footcite[p.~34-35]{glaser_rational_2010} Both states therefore benefit not just from spying, but from being spied \emph{upon}; their adversary's intelligence made that adversary more likely to propose or accept a risky signal for deescalation.

@@

\subsection{Assumptions and alternate explanations}
The analysis in this section relies on one crucial assumption: that the states in question are security-seekers. Analyses of the risks of espionage must contend with possibility that one of the states in question is \enquote{fundamentally dissatisfied with the status quo, desiring additional territory even when it is not required for security}---a greedy state.\footcite[p.~4]{glaser_rational_2010} Such a state would obviously want to protect both its motives and its plans, and therefore derive no benefit from having espionage conducted against it. Nonetheless, we know that essentially every major nation-state abides by the accepted norms of espionage. I have already addressed the possibility that states---including greedy states---are bound to these norms for reasons that extend beyond rational calculus, and do not consider that explanation to be sufficient. To fully evaluate the implications of greedy states on this theory of mutual cooperation is beyond the scope of this paper, but I will propose potential answers to the question of why greedy states would tolerate espionage as a starting point for future analysis.

Perhaps greedy states simply do not exist. This is a very provocative assertion---one that is, in its most stringent sense, almost certainly wrong---but consider the possibility that no states which abide by espionage norms are greedy \emph{to a destabilizing degree.} The concept of a greedy state is one that naturally operates on a spectrum.\footcite[p.~39. Throughout his book, Glaser simplifies and complicates his theory as is necessary for the particular analysis he is performing. Greedy states can be understood as the opposite of security-seeking states, or it can be its own independant variable, allowing for the possibility of states that are greedy \emph{and} security-seeking, or just one of the two.]{glaser_rational_2010} The quintessential \enquote{greediest} state, 1937 Germany, was planning to dominate the European continent. In 2020, if no states are worried about espionage revealing that they intend to launch a significant campaign of conquest, perhaps that is because there are no states intending to launch a significant campaign of conquest. To use a modern example, the CIA warned that Russia might try to move into the Crimean peninsula shortly before it happened in 2014.\footcite{hosenball_ukraine_2014} Russia practiced good operational security to protect its intent from espionage, but it still took a risk that its operation would be discovered and counteracted. If in fact the annexation of Crimea is the act of a greedy state, that level of greed is likely tolerable to Russia's adversaries in a way that a nuclear first-strike operation would not.

@@

\subsection{Aerial espionage}
Just a few weeks before the \enquote{Four Powers Summit} in Summer 1960, Soviet Premeir Nikita Khrushchev announced that the USSR had shot down an American plane.\footcite[p.~112]{powers_operation_2004} The CIA knew tgat we was referring to the U-2, an advanced spy plane that had bedeviled the USSR for years, proving beyond the reach of Soviet surface-to-air missiles.\footcite{orlov_u-2_2007} Nevertheless, the CIA was still confident that its cover story would hold, thanks to a morbid fact of the U-2's design: in the event of a shootdown, it should have been impossible for the pilot to survive.\footcite[p.~35]{lindgren_trust_2000} When the United States put out at a statement that the plane was part of a \enquote{weather reconnaissance} mission gone awry, Khrushchev sprung his trap---he revealed that the U-2 pilot, Gary Powers, had been taken alive.

@@

\subsubsection{Military reconnaissance flights}
Beginning in 1947, the US began a campaign of reconnaissance flights that routinely violated Soviet airspace. Converted Air Force bombers---called \enquote{ferrets}---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} The pilots who flew these missions were conducting one of the earliest forms of Cold War peacetime espionage: gathering SIGINT (signals intelligence) that identified the location of critical radar installations along the Soviet border. Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace.

@@

% The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}

Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted.\footnote{Some of the necessary information was so highly classified that it is likely lost forever.} Nonetheless, the civilian researcher is aided by the declassified research of military and intelligence archivists. Writing for the internal CIA publication \emph{Cryptologic Quarterly}, Michael Peterson compiled a now-declassified list of American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).\footnote{Michael Peterson is a former intelligence analyst and section chief for the CIA.} Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) involved US aircraft exclusively performing reconnaissance.\footcite[p.~4. Regardless of how thorough declassified CIA information appears, one cannot rule out the possibility that there were other, even more secret missions which, for whatever reason, the United States government has chosen to conceal to this day. Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article suggest that the information is relatively complete. First, the article was written for a classified audience, likely given that designation so that it could be comprehensive, and declassified after 16 years. Additionally, it appears to have been declassified without redactions or missing pages. Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.]{peterson_maybe_1993} In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5. In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea. Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this must be a typographical error, and corrected it here.]{peterson_maybe_1993}

@@

The Baltic incident exemplifies the reduced consequences for espionage that $H_2$ predicts, in both directions. The Soviet Union opened fire on a US Navy aircraft, which based on all available evidence did not violate the traditional 12-mile airspace boundary.\footcite[p.~7. To be extra specific]{peterson_maybe_1993} Given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite{the_new_york_times_soviet_1950} Under other circumstances, this could have been a major incident, but the United States chose not to publicize it. The Soviet Union, for its part, chose not to escalate the issue either, despite the ample evidence that the US was engaging in military reconnaissance.

The resolution of the Baltic Incident set a precedent for reconnaissance flights. From then on until 1960, the United States lost nine more aircraft to Soviet air defenses under similar circumstances. Each time, the pattern of responses was the same---each side released dueling statements of protest, then the situation continued more or less as before. A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later. This mission is listed by Peterson as a ferret flight, and Farquhar notes that the media thought the attack was intended to lower American prestige during election season, but does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website. As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions \enquote{of which little is known.}\footcite[p.~45]{farquhar_aerial_2015}

In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight, per Peterson) over the Sea of Japan. The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane. Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} After the September 4, 1954 shootdown of a Navy P2V Neptune, a US Senator publicly denounced the continued anodyne response: \enquote{just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, President Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets. A further signal of deescalation, the UN stunt demonstrated Eisenhower's ability to resist calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015} By late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking down.\footcite[p.~49]{farquhar_aerial_2015} In terms of both media coverage and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.

>>>>

\maketitle
\newpage
\input{abstract.tex}
\newpage
\section{Introduction}
In 2015, federal authorities discovered a massive data breach inside the United States Office of Personnel Management (OPM). Hackers had, over the course of months, downloaded reams of sensitive data, including a cache of applications for federal security clearance. These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment. \enquote{[I]f you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States. Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} In the case of the OPM hack, a vivid imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018. See \cite{sanger_trump_2018}}

@@

\subsection{Defining espionage norms}
There are few extant definitions of peacetime espionage, and no satisfying ones. The problem that aspiring espionage scholars inevitably encounter is that \enquote{peacetime espionage} is universally understood as a term of art but almost never defined by legal codes, or interpreted by case law. Most scholars who tackle this question settle on a definition like the one proposed by I\~{n}aki Navarrete and Russell Buchan: \enquote{\enquote{Peacetime espionage} is a colloquial rather than a legal term [used] to describe different methods of collecting confidential information from closed as opposed to open sources \textelp{} generally conducted by States, against States.}\footcite[901-902]{navarrete_out_2019} This definition purposefully excludes forms of extralegal government intervention that are often conflated with espionage, such as assassinations, election interference, and covert operations.\footnote{Often papers won't take a stand on defining espionage, but rather offer a working definition just to bound the scope of that particular paper. See \cite[600]{radsan_unresolved_2007}, \cite[2]{kapp_spying_2007})} Excluding these activities is important: assassinations and similar covert operations are not considered espionage, so they do not typically receive the protections of espionage norms.

@@

\subsection{Extant explanations for espionage norms}
Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage. In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments. See \cite{wright_essays_1962}} Espionage is universally illegal but every state acknowledges that allies and adversaries alike are guaranteed to engage in it.\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} Because of that tension, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact. Jared Beim concludes that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so there ought to be stronger enforcement mechanisms in place.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that at best \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[672]{beim_enforcing_2018} A. John Radsan employs a more philosophical approach. Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage generally reach the same conclusion: it would be almost impossible to argue that espionage is legal, but the international system leaves this tension in a state of salutary neglect. Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.

Another branch of scholarship sidesteps the legal question in favor of a \enquote{functional theory} of IR. This approach is more informative because it attempts to explain why states tolerate espionage rather than creating \emph{post hoc} justifications for it. A functional approach, for instance, might highlight the role of espionage in treaty verification, the way Christopher Baker does when he argues \enquote{the advantages that espionage offers over legally-binding verification and assurance regimes tip the scales in favor of functional cooperation.}\footcite{baker_tolerance_2004} Baker correctly notes that the additional assurance provided by espionage enables states to enter into riskier agreements that they would otherwise have avoided. His analysis is limited, however, by its focus on enabling specific policy outcomes, like treaties. It doesn't resolve the gap between the absolute illegality of espionage in domestic law and its widespread practice internationally, nor does address why espionage would be tolerated in cases where no tangible outcome, like a treaty, is forthcoming. Responding to Baker, Radsan notes that one-off deals and treaties \enquote{are a very slow and indirect way to achieve international consensus on the legality of espionage. The gap is still there.}\footcite[607]{radsan_unresolved_2007}

@@

Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage. The results of intelligence operations can have devastating, even existential effects on national security. In recent years China has relentlessly targeted American military secrets. A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} \enquote{Long-term, US future military advantage is being diminished by years of IP exfiltration from [US military agencies], all with little to no adverse consequences to the thieves.}\footcite[6]{bayer_cybersecurity_2019} The forecasts were dire: \enquote{If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.}\footcite[5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and we should be cautious about its conclusions; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see: \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states to gain a comparative advantage and the consequences for taking advantage of them remain minimal.

@@

\subsection{A model for maintaining espionage norms}
I propose an iterated decision-making model to describe how states cooperate to preserve the liminal legal status of espionage, based on the same tit-for-tat strategy associated with diplomatic expulsions. In game theory, tit-for-tat is highly-successful strategy wherein the player simply mimics the choice that that their opponent made in the last round.\footcite[8]{axelrod_effective_1980} By employing this strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same. Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind. Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal removal of diplomats to stop.

@@

To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure. Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[7]{glaser_rational_2010} If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond by not building up arms, and resolve the security dilemma for both of them. The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky. The state could be wrong that its adversary is a security seeker---what Glaser calls a \enquote{greedy state}---in which case its optimistic overtures of cooperation will have exposed it to greater risk of attack. When states are unwilling to make those signaling measures because they fear compromising their own security, then a security dilemma results.

Espionage creates the conditions that make solving the security dilemma possible. Consider a situation in which one state is a genuine security-seeker and so is its adversary. The state conducts espionage on its adversary, and the resulting intelligence increases its internal estimate of how likely the adversary is to be a security-seeker. As a result, the state decides that some form of arms control is no longer too risky to attempt in order to signal its benign intentions. Because the state receiving the signal is also a security-seeker, this is a welcome development; it could be empowered to signal its benign intentions as well, based on its own intelligence. For both states, this is the optimal outcome, contributing to both information variables that Glaser identifies as key to the security dilemma: \enquote{the state’s estimate of the adversary’s motives, and the state’s estimate of the adversary’s estimate of the state’s motives.}\footcite[34-35]{glaser_rational_2010} Both states therefore benefit not just from spying, but from being spied \emph{upon}; their adversary's intelligence made that adversary more likely to propose or accept a risky signal for deescalation.

@@

\subsection{Assumptions and alternate explanations}
The analysis in this section relies on one crucial assumption: that the states in question are security-seekers. Analyses of the risks of espionage must contend with possibility that one of the states in question is \enquote{fundamentally dissatisfied with the status quo, desiring additional territory even when it is not required for security}---a greedy state.\footcite[4]{glaser_rational_2010} Such a state would obviously want to protect both its motives and its plans, and therefore derive no benefit from having espionage conducted against it. Nonetheless, we know that essentially every major nation-state abides by the accepted norms of espionage. I have already addressed the possibility that states---including greedy states---are bound to these norms for reasons that extend beyond rational calculus, and do not consider that explanation to be sufficient. To fully evaluate the implications of greedy states on this theory of mutual cooperation is beyond the scope of this paper, but I will propose potential answers to the question of why greedy states would tolerate espionage as a starting point for future analysis.

Perhaps greedy states simply do not exist. This is a very provocative assertion---one that is, in its most stringent sense, almost certainly wrong---but consider the possibility that no states which abide by espionage norms are greedy \emph{to a destabilizing degree.} The concept of a greedy state is one that naturally operates on a spectrum.\footcite[39. Throughout his book, Glaser simplifies and complicates his theory as is necessary for the particular analysis he is performing. Greedy states can be understood as the opposite of security-seeking states, or it can be its own independant variable, allowing for the possibility of states that are greedy \emph{and} security-seeking, or just one of the two.]{glaser_rational_2010} The quintessential \enquote{greediest} state, 1937 Germany, was planning to dominate the European continent. In 2020, if no states are worried about espionage revealing that they intend to launch a significant campaign of conquest, perhaps that is because there are no states intending to launch a significant campaign of conquest. To use a modern example, the CIA warned that Russia might try to move into the Crimean peninsula shortly before it happened in 2014.\footcite{hosenball_ukraine_2014} Russia practiced good operational security to protect its intent from espionage, but it still took a risk that its operation would be discovered and counteracted. If in fact the annexation of Crimea is the act of a greedy state, that level of greed is likely tolerable to Russia's adversaries in a way that a nuclear first-strike operation would not.

@@

\subsection{Aerial espionage}
Just a few weeks before the \enquote{Four Powers Summit} in Summer 1960, Soviet Premeir Nikita Khrushchev announced that the USSR had shot down an American plane.\footcite[112]{powers_operation_2004} The CIA knew tgat we was referring to the U-2, an advanced spy plane that had bedeviled the USSR for years, proving beyond the reach of Soviet surface-to-air missiles.\footcite{orlov_u-2_2007} Nevertheless, the CIA was still confident that its cover story would hold, thanks to a morbid fact of the U-2's design: in the event of a shootdown, it should have been impossible for the pilot to survive.\footcite[35]{lindgren_trust_2000} When the United States put out at a statement that the plane was part of a \enquote{weather reconnaissance} mission gone awry, Khrushchev sprung his trap---he revealed that the U-2 pilot, Gary Powers, had been taken alive.

@@

\subsubsection{Military reconnaissance flights}
Beginning in 1947, the US began a campaign of reconnaissance flights that routinely violated Soviet airspace. Converted Air Force bombers---called \enquote{ferrets}---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[4]{peterson_maybe_1993} The pilots who flew these missions were conducting one of the earliest forms of Cold War peacetime espionage: gathering SIGINT (signals intelligence) that identified the location of critical radar installations along the Soviet border. Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace.

@@

% The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[8]{peebles_shadow_2000}

Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted.\footnote{Some of the necessary information was so highly classified that it is likely lost forever.} Nonetheless, the civilian researcher is aided by the declassified research of military and intelligence archivists. Writing for the internal CIA publication \emph{Cryptologic Quarterly}, Michael Peterson compiled a now-declassified list of American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).\footnote{Michael Peterson is a former intelligence analyst and section chief for the CIA.} Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) involved US aircraft exclusively performing reconnaissance.\footcite[4. Regardless of how thorough declassified CIA information appears, one cannot rule out the possibility that there were other, even more secret missions which, for whatever reason, the United States government has chosen to conceal to this day. Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article suggest that the information is relatively complete. First, the article was written for a classified audience, likely given that designation so that it could be comprehensive, and declassified after 16 years. Additionally, it appears to have been declassified without redactions or missing pages. Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.]{peterson_maybe_1993} In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[5. In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea. Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this must be a typographical error, and corrected it here.]{peterson_maybe_1993}

@@

The Baltic incident exemplifies the reduced consequences for espionage that $H_2$ predicts, in both directions. The Soviet Union opened fire on a US Navy aircraft, which based on all available evidence did not violate the traditional 12-mile airspace boundary.\footcite[7. To be extra specific]{peterson_maybe_1993} Given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite{the_new_york_times_soviet_1950} Under other circumstances, this could have been a major incident, but the United States chose not to publicize it. The Soviet Union, for its part, chose not to escalate the issue either, despite the ample evidence that the US was engaging in military reconnaissance.

The resolution of the Baltic Incident set a precedent for reconnaissance flights. From then on until 1960, the United States lost nine more aircraft to Soviet air defenses under similar circumstances. Each time, the pattern of responses was the same---each side released dueling statements of protest, then the situation continued more or less as before. A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later. This mission is listed by Peterson as a ferret flight, and Farquhar notes that the media thought the attack was intended to lower American prestige during election season, but does not mention any diplomatic consequences.\footcite[43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website. As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions \enquote{of which little is known.}\footcite[45]{farquhar_aerial_2015}

In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight, per Peterson) over the Sea of Japan. The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane. Despite press outrage, no further action was taken.\footcite[47]{farquhar_aerial_2015} After the September 4, 1954 shootdown of a Navy P2V Neptune, a US Senator publicly denounced the continued anodyne response: \enquote{just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers.}\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, President Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets. A further signal of deescalation, the UN stunt demonstrated Eisenhower's ability to resist calls for harsher action while still placating the American public.\footcite[47]{farquhar_aerial_2015} By late 1954, the \enquote{cycle of hostility} over aerial reconnaissance was already breaking down.\footcite[49]{farquhar_aerial_2015} In terms of both media coverage and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.
